---
title: "Height and Weight"
author: "David A Hughes"
date: "2026-01-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(patchwork)
library(cli)

source("function/extract_formula.R")
```


## read in the data

```{r}
f = "../data/processed/nhanes_combined_data.csv"
mydata = read_csv(f, show_col_types = FALSE)

## Edits sex
mydata = mydata |> mutate(sex = ifelse(sex == 1, "m", "f") )
mydata = mydata |> mutate(sex = as.factor(sex))

cli::cli_alert_info("Total individuals in dataset: {nrow(mydata)}")
```


## limit to adults / exclude kids

```{r}
adults = mydata |> filter(age >= 21) 
cli::cli_alert_info("Total adults: {nrow(adults)}")
```

## Male Female Count

```{r}
cli::cli_alert_info("Number of females and males:")
table(adults$sex)
```


## Ancestry / ethnicity composition

```{r}
cli::cli_alert_info("Ancestry composition:")
table(adults$ethnicity)
```

### ethnicity codes 

    - 1 = Mexican american, 
    - 2 = other Hispanic, 
    - 3 = non-Hispanic white, 
    - 4 = non-Hispanic black, 
    - 5 = other/multiracial



## edit height from cm to m

```{r}
adults = adults |> mutate(height = height * 0.01)
```


## estimate w/h and w/h^2 (aka BMI) and height squared

```{r}
adults = adults |> mutate(wh = weight/height, wh2 = weight/height^2, height2 = height^2)
```


## estimate log10(height) and log10(weight)

```{r}
adults = adults |> mutate( log10_height = log10(height), log10_weight = log10(weight) )

```


## estimate sex-centered log10(height) and log10(weight)

```{r}
res_w <- residuals(lm(log10_weight ~ sex, data = adults))
res_h <- residuals(lm(log10_height ~ sex, data = adults))

# adults = adults |> mutate(log10_height_sc = res_h + mean(adults$log10_height),
#                           log10_weight_sc = res_w + mean(adults$log10_weight) )

adults = adults |> mutate(log10_height_sc = res_h ,
                          log10_weight_sc = res_w  )

```


## log-log regressions

### Adults no adjustment for sex

```{r}
fit0 = lm( log10_weight ~ log10_height, data = adults )
cli::cli_alert_info("Model R squared: {round(summary(fit0)$r.squared, d = 4)}")

cli::cli_alert_info("Model formula")
extract_formula(fit0)
```

### Males only

```{r}
fitm = lm( log10_weight ~ log10_height, data = adults |> filter(sex == "m") )
cli::cli_alert_info("Model R squared: {round(summary(fitm)$r.squared, d = 4)}")

cli::cli_alert_info("Model formula")
extract_formula(fitm)
```

### Females only

```{r}
fitf = lm( log10_weight ~ log10_height, data = adults |> filter(sex == "f") )
cli::cli_alert_info("Model R squared: {round(summary(fitf)$r.squared, d = 4)}")

cli::cli_alert_info("Model formula")

extract_formula(fitf)
```

### Sex adjusted adults

```{r}
fit = lm( log10_weight ~ sex + log10_height, data = adults )
ss = anova(fit)[,2]
eta_sq = ss / sum(ss)
names(eta_sq) = rownames(anova(fit))

height_eta_sq = eta_sq["log10_height"]

cli::cli_alert_info("Model R squared: {round( summary(fit)$r.sq, d = 4)}")
cli::cli_alert_info("Partial R squared for height: {round(height_eta_sq, d = 4)}")

cli::cli_alert_info("Model formula")
extract_formula(fit)
```

## weight-height allometric model

The multiplicative allometric model (of the log-log regression) is:

```{r}
## interecept
intercept = summary(fit)$coefficients["(Intercept)", "Estimate"]
alpha = round(10^intercept, d = 3)

## sex coef
sex_coef = summary(fit)$coefficients["sexm", "Estimate"]
k = round(10^sex_coef, d = 3)

## height coef also called beta
beta = round( summary(fit)$coefficients["log10_height", "Estimate"] , d = 3)

## back converted, multiplicative
fit_multiplicative = paste0("weight = ", alpha, " * (", k, ")^sex * height^", beta )
fit_multiplicative
```




## plot log-log regression with formula and R squared for height


```{r}
f = gsub("0\\s*\\+\\s*", "", extract_formula(fit))
f = gsub("_", " ", f)
f = gsub("  ", " ", f)

rsq = paste0( "height partial R² = ",  round( height_eta_sq , d = 4) )


F1_A = adults |> ggplot( aes(x = log10_height, y = log10_weight ) ) +
  geom_point( alpha = 0.3, color = "grey40", size = 1.25 ) +
  geom_smooth( method = "lm", color = "black", se = TRUE, formula = 'y ~ x' ) +
  theme_bw() +
  labs(x = "log10 height (m)", y = "log10 weight (kg)", 
       #title = "Continuous NHANES adults (21+ years)",
       subtitle = "log-log allometric model") +
  annotate("text", x = 0.22, y = 1.48, label = f , size = 3.5, color = "black" ) +
  annotate("text", x = 0.22, y = 1.42, label = rsq , size = 3.5, color = "black" )
```

## estimate prediction line given log-log regression for W vs H plot

```{r}
## Prediction : DEFINED ABOVE
# alpha <- 27.932
# beta  <- 2.0492
# k <- 0.976  # multiplier when sex == "Male" (assuming female is reference)

newdat <- expand.grid(
  height = seq(min(adults$height, na.rm = TRUE),
               max(adults$height, na.rm = TRUE),
               length.out = 200),
  sex = levels(adults$sex)
)

newdat$pred_W <- alpha * (newdat$height^beta) * ifelse(newdat$sex == "Male", k, 1)
```



```{r}
F1_B = adults |> ggplot( aes(x = height, y = weight ) ) +
  geom_point( alpha = 0.3, color = "grey40", size = 1.25 ) +
  # geom_smooth( method = "lm", color = "blue", se = TRUE, formula = 'y ~ x' ) +
  geom_line(data = newdat, aes(y = pred_W), linewidth = 1) +
  theme_bw() +
  labs(x = "height (m)", y = "weight (kg)", 
       # title = "Continous NHANES adults (21+ years)",
       subtitle = "weight-height allometric model") +
  annotate("text", x = 1.8, y = 25, label = fit_multiplicative , size = 3.5, color = "black" ) 
```


```{r, fig.width = 6, fig.height = 10 }
library(patchwork)
fig1 = (F1_A / F1_B ) + plot_annotation(tag_levels = "A")

pdf("../results/figure1.pdf", width = 6, height = 10)
fig1
dev.off()

fig1
```



## sex-centered loq-log regression

### Sex adjusted adults

```{r}
fitsc = lm( log10_weight ~ log10_height, data = adults |> 
              dplyr::select(log10_weight_sc, log10_height_sc) |>
              rename(log10_weight = log10_weight_sc,
                     log10_height = log10_height_sc)
            )

## estiamte R squared
rsq =  paste0( "R² = ",  round( summary(fitsc)$r.squared , d = 4) )
cli::cli_alert_info("Model R squared: {rsq}")

cli::cli_alert_info("Model formula")
f = extract_formula(fitsc)
```


NOTE: the beta estimate is, as expected, the same. 


### A plot of the sex centered log-log regression

```{r}
pdata = data.frame(log10_height = adults$log10_height_sc, 
                   log10_weight = adults$log10_weight_sc ) 

## remove the 0 intercept
f = gsub("0\\s*\\+\\s*", "", extract_formula(fitsc))
f = gsub("_", " ", f)

pdata |> ggplot( aes(x = log10_height, y = log10_weight ) ) +
  geom_point( alpha = 0.3, color = "grey40", size = 1.25 ) +
  geom_smooth( method = "lm", color = "black", se = TRUE, formula = 'y ~ x' ) +
  theme_bw() +
  labs(x = "log10 height (m)", y = "log10 weight (kg)", 
       title = "Continous NHANES adults (21+ years)", 
       subtitle = "height and weight were sex-centered prior to log-log regression") +
  annotate("text", x = 0.0325, y = -0.38, label = f , size = 4.5, color = "black" ) +
  annotate("text", x = 0.0325, y = -0.45, label = rsq , size = 4.5, color = "black" )
```


## If log10(BMI) associates with log10(height) then Beta differs significantly from 2. 

```{r}
cli::cli_alert_info("No sex adjustment")
fit_test0 = lm(log10(wh2) ~ log10(height), data = adults )
summary(fit_test0)$coefficients

cli::cli_alert_info("Sex adjustment")
fit_test = lm(log10(wh2) ~ sex + log10(height), data = adults )
a = anova(fit_test)
eta_sq = a[,2] / sum(a[,2])

summary(fit_test)$coefficients
```


## Does WSI with a beta of 2.05 associate with height ?

```{r}
adults = adults |> mutate( wsi = weight / height^2.05 )
fit_wsi = lm(log10(wsi) ~ sex + log10(height), data = adults)
a = anova(fit_wsi)
eta_sq = a[,2] / sum(a[,2])

summary(fit_wsi)$coefficients
```

## IS BMI associated with % body fat in adults?

```{r}
## Males
fit_bmi_fat = lm( percent_fat ~  wh2, data = adults |> filter(sex == "m"))
summary(fit_bmi_fat)
```

```{r}
## Males
fit_bmi_fat = lm( percent_fat ~  wh2, data = adults |> filter(sex == "f"))
summary(fit_bmi_fat)
```

